---
title: "Agricultural Production Statistics in New Zealand"
subtitle: "TidyTuesday Week XX"
author: "Aghata Charnobay"
date: "2026-02-17"
categories: [time series, line plot]
page-navigation: true
toc: true
toc-depth: 3
format:
  html:
    theme: flatly
execute: 
  warning: false
  message: false
---

## A look at this week's dataset

This week we are exploring agriculture production statistics in New Zealand using data compiled from from StatsNZ.

## 1 Setup

### 1.1 Load libraries

```{r}
#| label: libraries

library(tidytuesdayR)
library(tidyverse)
library(tidytext)
library(ggtext)       
library(showtext) 
library(ggbranding)
library(stringr)
library(patchwork)

```


### 1.2 Set theme

```{r}
#| label: Theme and appearance

# Font setup 

font_add_google("Commissioner")
showtext_auto()
showtext_opts(dpi = 300)

# Theme style 

font_main  <- "Commissioner"
col_title  <- "grey10"
col_text   <- "grey30"
col_accent <- "#D26A44" 

crop_palette <- c(
  "Maize"  = "#ffca16",
  "Wheat"  = "#f97300",
  "Barley" = "#996515",
  "Oats"    = "#187322"
)

```

### 1.3 Load tidytuesday data

```{r}
#| label: load tidytuesday data

tuesdata <- tt_load(2026, week = 7)
dataset <- tuesdata$dataset

```


## 2 Quick exploratory data analysis

# The main grain crops grown in New Zealand are barley, maize, wheat, and oats.
# Let's check whether data for these crops are available in the dataset.

```{r}

#| label: EDA 1
dataset |> 
  count(measure, sort = TRUE) |>
  filter(str_detect(measure, "Wheat|Barley|Oats|Maize")) 

```

The dataset includes yield and area (sown/harvested) for these crops, although some values are missing.

 - Area sown: land planted.
 - Area harvested: land actually harvested.

Typically, area harvested is smaller than area sown due to losses.
However, in this dataset the terminology appears to change around 2003 from "area sown" to "area harvested." 

For consistency, I will treat them as equivalent measures of cultivated area.

## 3 Prepare data for plotting

```{r}
#| label: Prepare data for plotting

dataset_crops <- dataset |>
  filter(str_detect(measure, "Wheat|Barley|Oats|Maize")) |>
  mutate(
    crop = str_extract(measure, "Wheat|Barley|Oats|Maize"),
    variable = if_else(
      str_detect(str_to_lower(measure), "area"),
      "area_harvested",
      "yield"
    )
  ) |>
  select(year_ended_june, crop, variable, value) |>
  pivot_wider(
    names_from = variable,
    values_from = value
  ) |>
  mutate(
    productivity = yield / area_harvested,
    yield_kt = yield / 1000,
    area_harvested_thousand_ha = area_harvested / 1000
  )

```

## 4 Plot

```{r}
#| label: Plot data

# Plot 1: Area Harvested
p_area <- ggplot(dataset_crops,
                 aes(x = year_ended_june,
                     y = area_harvested_thousand_ha,
                     color = crop)) +
  geom_line(linewidth = 0.7, alpha = 0.7) +
  scale_color_manual(values = crop_palette) +
  labs(x = NULL, y = "Area (thousand ha)") +
  theme_minimal(base_family = font_main) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 8, color = "grey30"),
        axis.title = element_text(face = "bold", size = 9, color = "grey30"))

# Plot 2: Yield
p_yield <- ggplot(dataset_crops,
                 aes(x = year_ended_june,
                     y = yield_kt,
                     color = crop)) +
  geom_line(linewidth = 0.7, alpha = 0.7) +
  scale_color_manual(values = crop_palette) +
  labs(x = NULL, y = "Yield (thousand tonnes)") +
  theme_minimal(base_family = font_main) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 8, color = "grey30"),
        axis.title = element_text(face = "bold", size = 9, color = "grey30"))

# Plot 3: Productivity
p_productivity <- ggplot(dataset_crops,
                 aes(x = year_ended_june,
                     y = productivity,
                     color = crop)) +
  geom_line(linewidth = 0.7, alpha = 0.7) +
  scale_color_manual(values = crop_palette) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1)) +
  labs(x = "", y = "Productivity (tonnes/ha)") +
  theme_minimal(base_family = font_main) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 8, color = "grey30"),
        axis.title = element_text(face = "bold", size = 9, color = "grey30"))

# Plot 4: Annotation Tag
p_text <- ggplot() +
  ggtext::geom_textbox(
    aes(x = 0.5, y = 0.5), 
    label = "<b style='color:#996515;'>Barley</b>, <b style='color:#f97300;'>wheat</b>, 
    <b style='color:#ffca16;'>maize</b>, and <b style='color:#187322;'>oats</b> dominate New Zealand’s cereal production.
    Over time, crop productivity for these grains has increased with cultivated 
    area remaining relatively stable, indicating that growth in cereal production has been driven by 
    improvements in technology and crop management rather than land expansion.",
    hjust = 0.5, vjust = 0.5, 
    halign = 0.5,               
    fill = NA, 
    box.colour = NA, 
    family = font_main,
    size = 3.5,                   
    lineheight = 1.2,
    width = unit(0.95, "npc")   # Leaves a small margin so it doesn't touch the other plots
  ) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  theme_void()

# 5. Assembly using Patchwork Design

layout_design <- "
  AABB
  CCDD
"

final_plot <- (p_area + p_yield + p_productivity + p_text) + 
  plot_layout(design = layout_design) +
  plot_annotation(
    title = "The Evolution of Cereal Production in New Zealand",
    theme = theme(
      plot.title = element_text(family = font_main, face = "bold", size = 16, color = col_title, margin = margin(b = 25)),
      plot.subtitle = element_markdown(),
      plot.margin = margin(20, 20, 10, 20)
    )
  ) 

p <- final_plot +
  add_branding(
    github          = "a-charnobay",
    linkedin        = "Aghata Charnobay",
    additional_text = "Data: Agriculture StatsNZ | #TidyTuesday 2026-Week 7",
    text_position   = "before",
    line_spacing    = 1L,
    text_color      = col_text,
    text_family     = font_main,
    icon_color      = col_accent,
    caption_halign  = 1,
    icon_size       = "9pt",
    text_size       = "9pt",
    caption_margin  = margin(t = 5, b = 5, unit = "pt")
  )

```


## 5 Save plot

```{r}
#| label: Save plot
#| eval: false

ggsave(
  filename = "NZ_agriculturalStats.png", 
  plot = p,
  width = 8, 
  height = 5,
  dpi = 300,
  bg = "white"
)
```




