---
title: "Science Foundation Ireland Grants Commitments"
subtitle: "TidyTuesday Week 08"
author: "Aghata Charnobay"
date: "2026-02-23"
categories: [timeseries,bump chart]
page-navigation: true
toc: true
toc-depth: 3
format:
  html:
    theme: flatly
execute: 
  warning: false
  message: false
---

## A look at this week's dataset

This week we're exploring Irish STEM research and ancillary projects funded by Science Foundation Ireland (SFI) since its foundation in 2000 to its dissolution on 31st July 2024. The data comes from [Ireland's Open Data Portal](https://data.gov.ie/dataset/science-foundation-ireland-grants-commitments).

> SFI was the national foundation in Ireland for investment in scientific and engineering research. Consequently, SFI invested in those academic researchers and research teams who were most likely to generate new knowledge, leading edge technologies and competitive enterprises in the fields of science, technology, engineering and maths (STEM).

## 1 Setup

### 1.1 Load libraries

```{r}
#| label: libraries

library(tidytuesdayR)
library(tidyverse)
library(lubridate)
library(ggbump)
library(ggtext)
library(showtext)
library(here)

```

### 1.2 Set theme

```{r}
#| label: Theme and appearance
#| 
font_add_google("Commissioner")
showtext_auto()
font_main <- "Commissioner"

# Font Awesome for caption
font_add(family = "fa-brands", regular = here("fonts", "Font Awesome 7 Brands-Regular-400.otf"))

# Colors
title_col <- "grey10"
text_col  <- "grey30"
bg_col    <- "#F2F4F8"

# Highlight colors for plot
highlight_colors <- c(
  "UCD"     = "#1F4E79",
  "TCD"     = "#0097A7",
  "Teagasc" = "#43A047",
  "Other"   = "grey90",
  "grey20"  = "grey20"
)
```

### 1.3 Load tidytuesday data

```{r}
#| label: load tidytuesday data

tuesdata <- tt_load(2026, week = 8)
sfi_grants <- tuesdata$sfi_grants

# Preprocess: dates, year, duration
sfi_grants_clean <- sfi_grants %>%
  mutate(
    start_date = as.Date(start_date),
    end_date   = as.Date(end_date),
    year       = year(start_date),
    grant_duration_days  = as.numeric(end_date - start_date),
    grant_duration_years = grant_duration_days / 365
  )

```

## 2 Quick exploratory data analysis

### 2.1 Total funding and number of grants over years

```{r}
#| label: EDA 1

sfi_summary <- sfi_grants_clean %>%
  filter(year >= 2000, year <= 2024) %>%
  group_by(year) %>%
  summarise(
    total_funding = sum(current_total_commitment, na.rm = TRUE),
    n_grants      = n(),
    avg_grant     = total_funding / n_grants,
    .groups = "drop"
  )

# Total funding over time
ggplot(sfi_summary) +
  geom_line(aes(x = year, y = total_funding), size = 1.2) +
  labs(title = "Total Funding Over time", x = "Year", y = "") +
  theme_minimal(base_family = font_main)

# Number of grants over time
ggplot(sfi_summary) +
  geom_line(aes(x = year, y = n_grants), size = 1.2) +
  labs(title = "Number of Grants Over time", x = "Year", y = "") +
  theme_minimal(base_family = font_main)


```

### 2.2 Distribution of grant durations

```{r}

#| label: EDA 2

ggplot(sfi_grants_clean, aes(x = grant_duration_years)) +
  geom_histogram(binwidth = 0.5, fill = "grey40", color = "white") +
  scale_x_continuous(breaks = seq(0, 15, 1)) +
  labs(title = "Distribution of Grant Durations in Years", x = "Duration (Years)", y = "Number of grants") +
  theme_minimal(base_family = font_main)

```

## 3 Prepare data for plotting

```{r}

#| label: Prepare data for plotting

# Top 10 institutions 2020-2024
top10 <- sfi_grants_clean %>%
  filter(year >= 2020, year <= 2024) %>%
  group_by(research_body) %>%
  summarise(total_funding = sum(current_total_commitment, na.rm = TRUE), .groups = "drop") %>%
  slice_max(total_funding, n = 10) %>%
  pull(research_body)

# Ranking data for bump chart
rank_data <- sfi_grants_clean %>%
  filter(year >= 2020, year <= 2024, research_body %in% top10) %>%
  group_by(year, research_body) %>%
  summarise(total_funding = sum(current_total_commitment, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  mutate(rank = dense_rank(desc(total_funding))) %>%
  ungroup() %>%
  mutate(
    research_body = str_remove(research_body, "^RCSI "),
    label = case_when(
      str_detect(research_body, "UCD")      ~ "UCD",
      str_detect(research_body, "TCD")      ~ "TCD",
      str_detect(research_body, "Galway")   ~ "Galway",
      str_detect(research_body, "UL")       ~ "UL",
      str_detect(research_body, "UCC")      ~ "UCC",
      str_detect(research_body, "RCSI")     ~ "RCSI",
      str_detect(research_body, "TNI")      ~ "TNI",
      str_detect(research_body, "Teagasc")  ~ "Teagasc",
      str_detect(research_body, "DCU")      ~ "DCU",
      str_detect(research_body, "MU")       ~ "MU",
      TRUE ~ research_body
    ),
    highlight = if_else(label %in% c("UCD", "TCD", "Teagasc"), label, "Other")
  )

```

## 4 Plot

```{r}

#| label: Plot data
#| fig-height: 7
#| fig-width: 4


p <- ggplot(rank_data, aes(x = year, y = rank, group = research_body)) +
  geom_bump(aes(color = highlight, size = highlight != "Other"), smooth = 8) +
  geom_point(aes(color = highlight), size = 2.5) +
  geom_text(
    data = rank_data %>% filter(year == 2024) %>% arrange(rank),
    aes(label = research_body, color = if_else(highlight == "Other", "grey20", highlight)),
    hjust = 0, nudge_x = 0.08, size = 2, show.legend = FALSE
  ) +
  scale_color_manual(values = highlight_colors) +
  scale_size_manual(values = c(`TRUE` = 1.4, `FALSE` = 0.7), guide = "none") +
  scale_y_reverse(breaks = 1:10) +
  scale_x_continuous(breaks = 2020:2024, expand = expansion(mult = c(0.02, 0.4))) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Who Leads Research Funding in Ireland?",
    subtitle = "Ranking of institutions based on the total research grants (€) awarded by Science Foundation Ireland (SFI)",
    caption = paste0(
      "**Data**: Ireland's Open Data Portal - TidyTuesday 2026, Week 08",
      "<br>**Design**: <span style='font-family:fa-brands; color:grey30;'>\uf09b</span> a-charnobay ",
      "<span style='font-family:fa-brands; color:grey30;'>\uf08c</span> Aghata Charnobay"
    ),
    x = NULL, y = NULL
  ) +
  theme_minimal(base_family = font_main) +
  theme(
    plot.margin = margin(10, 30, 5, 30),
    plot.title = element_text(face = "bold", size = 14, color = title_col, hjust = 0.5),
    plot.subtitle = element_text(size = 8, face = "italic", color = text_col, hjust = 0.5),
    plot.caption = element_markdown(size = 6, color = text_col, hjust = 0.5),
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 7)
  )
p
```

## 4.1 Why?

```{r}
#| label: The why behind Teagasc 2024 rise

# Analyze why Teagasc funding spiked in 2024

teagasc_2024 <- sfi_grants_clean %>%
  filter(research_body == "Teagasc", year == 2024)

# Summary stats
teagasc_summary <- teagasc_2024 %>%
  summarise(
    total_funding = sum(current_total_commitment, na.rm = TRUE),
    n_grants      = n(),
    avg_grant     = mean(current_total_commitment, na.rm = TRUE),
    .groups = "drop"
  )

teagasc_summary

# Top 10 grants
top_projects <- teagasc_2024 %>%
  arrange(desc(current_total_commitment)) %>%
  select(proposal_title, programme_name, current_total_commitment) %>%
  slice_head(n = 10)

top_projects

# Share of total funding
teagasc_shares <- teagasc_2024 %>%
  mutate(share = current_total_commitment / sum(current_total_commitment, na.rm = TRUE)) %>%
  arrange(desc(share))

teagasc_shares
```

## 5 Save plot

```{r}
#| label: Save plot
#| eval: false

ggsave(
  filename = "research_funding_ireland.png", 
  plot = p,
  width = 7, 
  height = 4,
  units = "in",
  dpi = 300,
  bg = "white"
)
```







