---
title: "One Million Digits of Pi"
subtitle: "TidyTuesday Week 12"
author: "Aghata Charnobay"
date: "2026-03-24"
categories: [Generative Art]
page-navigation: true
toc: true
toc-depth: 3
format:
  html:
    theme: flatly
execute: 
  warning: false
  message: false
---

## A look at this week's dataset

This week we're exploring the digits of π with a dataset submission in celebration of Pi Day (March 14). The dataset contains the first one million digits of π, beginning with 3.14159 …, collected from publicly available sources and curated for analysis and visualization.

> "Pi is an irrational number, meaning its decimal representation never ends and never settles into a permanent repeating pattern."" — *Eve Andersson, collector of the one million digits dataset*.

## 1 Setup

### 1.1 Load libraries

```{r}
#| label: libraries

library(tidytuesdayR)
library(here)
library(tidyverse)
library(tidytext)
library(ggtext)       
library(showtext) 
library(stringr)
library(lubridate)

```

### 1.2 Set theme

```{r}
#| label: Theme and appearance

# Font setup 
font_add_google("Caveat")
showtext_auto()
showtext_opts(dpi = 300)
font_main <- "Caveat"

# Font Awesome for caption
font_add(family = "fa-brands", regular = here("fonts", "Font Awesome 7 Brands-Regular-400.otf"))

# Colors
title_col <- "grey10"
text_col  <- "grey30"
bg_col    <- "#F2F4F8"

```

### 1.3 Load tidytuesday data

```{r}
#| label: load tidytuesday data

tuesdata <- tt_load(2026, week = 12)
pi_digits <- tuesdata$pi_digits

```

## 2 Prepare data for plotting

```{r}
#| label: Prepare data for plotting


pi_palette <- c(
  "#0B3D2E", # green
  "#CFE3D6", # green-light
  "#06261B", # green-dark
  "#D97A2B", # orange
  "#FBE6C8", # orange-light
  "#A8561E", # orange-dark
  "#4A2C1D", # brown-dark
  "#9C7662", # brown-warm
  "#4E6A78", # blue-gray
  "#F5F5DC"  # tan-light
)

grid_size <- 100 

plot_data <- pi_digits %>%
  filter(digit_position <= grid_size^2) %>%
  mutate(
    x = (digit_position - 1) %% grid_size,
    y = desc((digit_position - 1) %/% grid_size),
    digit = as.factor(digit)
  )

```

## 3 Plot

```{r}
#| label: Plot data


p1<- ggplot(plot_data, aes(x = x, y = y, color = digit)) +
  geom_point(shape = 19, size = 1.5) + 
  scale_color_manual(values = pi_palette) +
  labs(caption = paste0(
      "**Data**: PiDay - TidyTuesday 2026, Week 12",
      "<br>**Design**: <span style='font-family:fa-brands; color:#D97A2B;'>&#xf09b;</span> a-charnobay ",
      "<span style='font-family:fa-brands; color:#D97A2B;'>&#xf08c;</span> Aghata Charnobay"
    ))+
  theme_minimal(base_family = font_main) +
  theme(
    plot.background = element_rect(fill = "#F8F9FA"),
    plot.caption = element_markdown(size = 10, color = text_col, margin = margin(t = 5), lineheight = 1.1, hjust = 0.5),
     legend.position = "none",
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank()
  )

p1

```

## 4 Save plot

```{r}
#| label: Save plot
#| eval: false

ggsave(
  filename = "pi_digits.png", 
  plot = p1,
  width = 6, 
  height = 6,
  dpi = 300,
  bg = "white"
)
```
