---
title: "2026 Winter Olympics"
subtitle: "TidyTuesday Week 06"
author: "Aghata Charnobay"
date: "2026-02-10"
categories: [distribution, heatmap]
page-navigation: true
toc: true
toc-depth: 3
format:
  html:
    theme: flatly
execute: 
  warning: false
  message: false
---

## A look at this week's dataset

This week we're exploring the event schedule for the 2026 Winter Olympics in Milan-Cortina, Italy. The dataset contains detailed information about all 1,866 Olympic events, including both competition and training sessions across various winter sport disciplines.

## 1 Setup

### 1.1 Load libraries

```{r}
#| label: libraries

library(tidytuesdayR)
library(tidyverse)
library(tidytext)
library(ggtext)       
library(showtext) 
library(ggbranding)
library(stringr)

```


### 1.2 Set theme

```{r}
#| label: Theme and appearance

# Font setup 

font_add_google("Commissioner")
showtext_auto()
showtext_opts(dpi = 300)

# Theme style 

font_main  <- "Commissioner"
col_title  <- "grey10"
col_text   <- "grey30"
col_accent <- "#926c15" 
color_palette <- c("#E5E5E5", "#E6D5A9", "#dbb42c", "#926c15")
```

### 1.3 Load tidytuesday data

```{r}
#| label: load tidytuesday data

tuesdata <- tt_load(2026, week = 6)
olympic_schedule <- tuesdata$schedule

```


## 2 Quick exploratory data analysis

### How many events exist per discipline?

```{r}

#| label: EDA 1
olympic_schedule |> count(discipline_name, sort = TRUE) 

```

### On which day and at what time do most medal events occur?

```{r}
#| label: EDA 2

olympic_schedule |>
  filter(is_medal_event == TRUE) |> 
  count(day_of_week, hour = hour(start_datetime_utc)) |> 
  arrange(desc(n))
```


## 3 Prepare data for plotting

```{r}
#| label: Prepare data for plotting

heatmap_data <- olympic_schedule |>
  filter(is_medal_event == TRUE) |> 
  mutate(
    hour = hour(start_datetime_utc),
    day_of_week = fct_relevel(day_of_week, 
                               "Sunday", "Saturday", "Friday", "Thursday", 
                               "Wednesday", "Tuesday", "Monday") 
  ) |>
  count(day_of_week, hour)
```

## 4 Plot

```{r}
#| label: Plot data

p <- ggplot(heatmap_data, aes(x = hour, y = day_of_week, fill = n)) +
  geom_tile(color = "white", size = 1.5) + 
  scale_fill_gradientn(colors = color_palette) +
  scale_x_continuous(breaks = seq(0, 23, 2), labels = paste0(seq(0, 23, 2), ":00")) +
  theme_minimal(base_family = font_main) +
  labs(
    title = "2026 Winter Olympics Golden Hour",
    subtitle = "Want to see the finals? Grab the remote on Tuesday, Wednesday, or the weekend — medal events hit their\npeak right after lunch!",
    x = "Hour (UTC timezone)",
    y = NULL
  ) +
  theme(
    plot.margin = margin(20, 20, 10, 20),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(face = "bold", size = 16, color = "black"),
    plot.title.position = "plot",
    plot.subtitle = element_text(size = 11, color = "grey30", margin = margin(t = 5, b = 20)),
    axis.text = element_text(face = "bold", size = 10, color = "grey30"),
    axis.title = element_text(size = 11, color = "black"),
    legend.position = "none"
  ) +
  add_branding(
    github                = "a-charnobay",
    linkedin              = "Aghata Charnobay",
    additional_text       = "Data: 2026 Winter Olympics Event Schedule | #TidyTuesday 2026 - Week 6",
    text_position         = "before",
    line_spacing          = 1L,
    text_color            = col_text,   
    text_family           = font_main,  
    icon_color            = col_accent, 
    caption_halign        = 1,
    icon_size             = "10pt",
    text_size             = "10pt",
    caption_margin        = margin(t = 25, b = 5, unit = "pt")
  )
p 
```


## 5 Save plot

```{r}
#| label: Save plot
#| eval: false

ggsave(
  filename = "winter_olympics.png", 
  plot = p,
  width = 8, 
  height = 5,
  dpi = 300,
  bg = "white"
)
```

