---
title: "Golem Grad Tortoise Data"
subtitle: "TidyTuesday Week 09"
author: "Aghata Charnobay"
date: "2026-03-03"
categories: [distribution, density]
page-navigation: true
toc: true
toc-depth: 3
format:
  html:
    theme: flatly
execute: 
  warning: false
  message: false
---

## A look at this week's dataset

he datasets this week come from the paper "Sex Ratio Bias Triggers Demographic Suicide in a Dense Tortoise Population". This topic seemed so interesting that even media like the New York Times picked it up: "Constant Sexual Aggression Drives Female Tortoises to Walk Off Cliffs".

> In an exceptionally dense island population of Hermann's tortoises in Lake Prespa in North Macedonia, sexually coercive males dramatically overnumber females, inflict severe copulatory injuries and put them at risk of fatal falls from the island plateau's sheer rock faces. Harassed females are emaciated, reproduce less frequently, produce smaller clutches and have lower annual survival rates compared to females from a neighbouring mainland population. Sixteen years of capture-recapture data reveal an ongoing extinction event and predict that the last island female will die in 2083.

## 1 Setup

### 1.1 Load libraries

```{r}
#| label: libraries

library(tidytuesdayR)
library(tidyverse)
library(tidytext)
library(ggtext)       
library(showtext) 
library(stringr)
library(here)
library(patchwork)
library(ggridges)

```


### 1.2 Set theme

```{r}
#| label: Theme and appearance
#| 
font_add_google("Commissioner")
showtext_auto()
showtext_opts(dpi = 300)
font_main <- "Commissioner"

# Font Awesome for caption
font_add(family = "fa-brands", regular = here("fonts", "Font Awesome 7 Brands-Regular-400.otf"))

# Colors
title_col <- "grey10"
text_col  <- "grey30"
bg_col    <- "#F2F4F8"

palette_color <- c(
  "Males"     = "#4E8644",
  "Females"     = "#ECB0B6"
  )

# Setting global theme defaults

theme_set(
  theme_minimal(base_family = font_main) +
    theme(
      plot.title = element_text(face = "bold", size = 16, color = title_col),
      plot.margin = margin(20, 20, 10, 20),
      legend.position = "none",
      strip.text = element_text(face = "bold")
    )
)

```

### 1.3 Load tidytuesday data

```{r}
#| label: load tidytuesday data

tuesdata <- tt_load(2026, week = 9)
clutch_size_cleaned <- tuesdata$clutch_size_cleaned
tortoise_body_condition_cleaned <- tuesdata$tortoise_body_condition_cleaned

```


## 2 Quick exploratory data analysis

### Do males and females differ in body condition?

```{r}
#| label: EDA 1
#| fig-width: 10
#| fig-height: 5

# pre-process
tortoise_body_condition_cleaned <- tortoise_body_condition_cleaned |>
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Females", "Males")),
    locality = factor(locality, levels = c("Konjsko", "Beach", "Plateau")),
    # fct_collapse to group levels and create the location factor at the same time
    location = fct_collapse(locality,
      "Mainland" = "Konjsko",
      "Island"   = c("Beach", "Plateau")
    )
  )

p1 <- ggplot(tortoise_body_condition_cleaned, aes(sex, body_mass_grams, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ location) +
  labs(x = "Sex", y = "Body Mass (g)")

p2 <- ggplot(tortoise_body_condition_cleaned, aes(sex, straight_carapace_length_mm, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ location) +
  labs(x = "Sex", y = "Carapace Length (mm)")

p3 <- ggplot(tortoise_body_condition_cleaned, aes(sex, body_condition_index, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ location) +
  labs(x = "Sex", y = "BCI")

(p1 + p2 + p3) + 
  plot_annotation(title = "Body Condition by Sex") & 
  scale_fill_manual(values = palette_color)

```

### How body condition changed over time

```{r}
#| label: EDA 2
#| fig-width: 8
#| fig-height: 8

stats_time <- tortoise_body_condition_cleaned |>
  group_by(year, sex, location) |>
  summarise(
    m_mass = mean(body_mass_grams, na.rm = TRUE),
    m_length = mean(straight_carapace_length_mm, na.rm = TRUE),
    m_bci = mean(body_condition_index, na.rm = TRUE),
    .groups = "drop"
  )

g1 <- ggplot(stats_time, 
             aes(year, m_mass, color = sex)) + 
  geom_line() + 
  geom_point() + 
  facet_wrap(~location) + 
  labs(y = "Mean Mass (g)", x= "")

g2 <- ggplot(stats_time, 
             aes(year, m_length, color = sex)) + 
  geom_line() + 
  geom_point() + 
  facet_wrap(~location) + 
  labs(y = "Mean Length (mm)", x= "")

g3 <- ggplot(stats_time, 
             aes(year, m_bci, color = sex)) + 
  geom_line() + 
  geom_point() + 
  facet_wrap(~location) + 
  labs(y = "Mean BCI", x = "Year")

(g1 / g2 / g3) + 
  plot_annotation(title = "Body Condition by Sex and Time") & 
  scale_color_manual(values = palette_color)

```

```{r}
#|label: EDA 3
#| fig-width: 10
#| fig-height: 5

loc_palette <- c("Konjsko" = "#88CCEE", "Beach" = "#CC6677", "Plateau" = "#DDCC77")

p1 <- ggplot(clutch_size_cleaned, aes(x = locality, y = eggs, fill = locality)) +
  geom_boxplot(alpha = 0.7) +
  labs(x = "Locality", y = "Clutch Size (Eggs)")

p2 <- ggplot(clutch_size_cleaned, aes(x = locality, y = body_mass_grams, fill = locality)) +
  geom_boxplot(alpha = 0.7) +
  labs(x = "Locality", y = "Body Mass (g)")

p3 <- ggplot(clutch_size_cleaned, aes(x = locality, y = straight_carapace_length_mm, fill = locality)) +
  geom_boxplot(alpha = 0.7) +
  labs(x = "Locality", y = "Carapace Length (mm)")

(p1 + p2 + p3) +
  plot_annotation(
    title = "Reproductive Traits by Locality",
    theme = theme(plot.title = element_text(size = 16, margin = margin(b = 20)))
  ) & 
  scale_fill_manual(values = loc_palette) &
  theme(legend.position = "none")          
```

## 3 Prepare data for plotting

```{r}
#| label: Prepare data for plotting

# Prepare data
female_data <- tortoise_body_condition_cleaned |>
  filter(sex == "Females") |>
  mutate(
    locality_label = factor(locality, 
                            levels = c("Plateau", "Beach", "Konjsko"),
                            labels = c("Plateau\n(Island)", "Beach\n(Island)", "Konjsko\n(Mainland)"))
  )

# Summary stats: No need for manual y_min/y_max if we use the factor levels
summary_stats <- female_data |>
  group_by(locality_label) |>
  summarise(
    n = n(),
    mean_bci = mean(body_condition_index, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    y = as.numeric(locality_label),
    label = paste0("n=", n, " | mean=", round(mean_bci, 1)),
    x_label = max(female_data$body_condition_index, na.rm = TRUE) + 3
  )
```

## 4 Plot

```{r}
#| label: Plot data

p<- ggplot(female_data, aes(x = body_condition_index, y = locality_label)) +
  geom_density_ridges(
    fill = palette_color["Females"], 
    alpha = 0.7,
    scale = 1.2,
    rel_min_height = 0.01,
    color = "black",
    ) +
  # Summary text (inherited from summary_stats)
  geom_text(
    data = summary_stats,
    aes(x = x_label, y = locality_label, label = label),
    nudge_y = 0.1, hjust = 1, size = 3.5, family = font_main, inherit.aes = FALSE
  ) +
  labs(
    title = "Mapping Tortoise Health Across Lake Prespa",
    subtitle = "The Body Condition Index (BCI) measures tortoise’s mass against its expected size. It's a key indicator<br>of health: higher scores represent better condition. Data from Lake Prespa (North Macedonia) shows<br>that <span style='color:#ECB0B6;'>**female**</span> island tortoises present lower BCI levels than those on the neighboring mainland.",
    x = "BCI", 
    y = NULL,
    caption = paste0(
      "**Data**: Arsovski et al. (2026) - TidyTuesday 2026, Week 09",
      "<br>**Design**: <span style='font-family:fa-brands; color:grey30;'>\uf09b</span> a-charnobay ",
      "<span style='font-family:fa-brands; color:grey30;'>\uf08c</span> Aghata Charnobay"
    )
  ) +
  theme(
    plot.title.position = "plot",
    plot.subtitle = element_markdown(size = 10, margin = margin(t=5, b=15), lineheight = 1.1),
    plot.caption = element_markdown(size = 9, color = text_col, margin = margin(t=15), lineheight = 1.1),
    panel.grid.major.x = element_blank()
  )
```


## 5 Save plot

```{r}
#| label: Save plot
#| eval: false

ggsave(
  filename = "tortoises.png", 
  plot = p,
  width = 7, 
  height = 5,
  dpi = 300, 
  bg = "white"
)
```
