---
title: "How likely is 'likely'?"
subtitle: "TidyTuesday Week 10"
author: "Aghata Charnobay"
date: "2026-10-03"
categories: [distribution, ridge]
page-navigation: true
toc: true
toc-depth: 3
format:
  html:
    theme: flatly
execute: 
  warning: false
  message: false
---

## A look at this week's dataset

In an [online quiz](https://probability.kucharski.io), created as an independent project by Adam Kucharski, over 5,000 participants compared pairs of probability phrases (e.g. “Which conveys a higher probability: Likely or Probable?”) and assigned numerical values (0–100%) to each of 19 phrases. The resulting data can be used to analyse how people interpret common probability phrases

## 1 Setup

### 1.1 Load libraries

```{r}
#| label: libraries

library(tidytuesdayR)
library(tidyverse)
library(tidytext)
library(ggtext)       
library(showtext) 
library(stringr)
library(here)
library(patchwork)
library(ggridges)

```

### 1.2 Set theme

```{r}
#| label: Theme and appearance
#| 
font_add_google("Commissioner")
showtext_auto()
showtext_opts(dpi = 300)
font_main <- "Commissioner"

# Font Awesome for caption
font_add(family = "fa-brands", regular = here("fonts", "Font Awesome 7 Brands-Regular-400.otf"))

# Colors
title_col <- "grey10"
text_col  <- "grey30"
bg_col    <- "#F2F4F8"



```

### 1.3 Load tidytuesday data

```{r}
#| label: load tidytuesday data

tuesdata <- tt_load(2026, week = 10)
absolute_judgements <- tuesdata$absolute_judgements
pairwise_comparisons <- tuesdata$pairwise_comparisons
respondent_metadata <- tuesdata$respondent_metadata

```

## 2 Prepare data for plotting

```{r}
#| label: Prepare data for plotting

ordered_terms <- c(
  "Almost No Chance", "Highly Unlikely", "Remote Chance", "Improbable", 
  "Chances are Slight", "Little Chance", "Unlikely", "Might Happen", 
  "Could Happen", "Realistic Possibility", "About Even", "May Happen", 
  "Better than Even", "Probable", "Likely", "Very Good Chance", 
  "Highly Likely", "Almost Certain", "Will Happen"
)

ridge_data <- absolute_judgements |>
  mutate(term = factor(term, levels = rev(ordered_terms)))

```

## 3 Plot

```{r}
#| label: Plot data

p <- ggplot(ridge_data, aes(x = probability, y = term, fill = ..x..)) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01
  ) +
  scale_fill_gradient(
    low = "#A37AD9",   # conservative
    high = "#2E7D5B"   # optimistic
  ) +
  labs(
    title = "What does “Likely” mean?",
    subtitle = "When someone says something is “likely” to happen, what chance do you think (from 0 to<br> 100%) it has of actually happening? Data show people tend to agree on extreme terms but<br>interpret middle-range words very differently.",
    x = "Assigned Numerical Probability (%)",
    y = "",
    caption = paste0(
      "**Data**: CAPphrase: Adam Kucharski - TidyTuesday 2026, Week 10",
      "<br>**Design**: <span style='font-family:fa-brands; color:#A37AD9;'>\uf09b</span> a-charnobay ",
      "<span style='font-family:fa-brands; color:#A37AD9;'>\uf08c</span> Aghata Charnobay"
    )
  ) +
  theme_minimal(base_family = font_main) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16,
      color = title_col
    ),
    plot.title.position = "plot",
    plot.margin = margin(20, 60, 10, 20),
    plot.subtitle = element_markdown(
      size = 11,
      margin = margin(t = 5, b = 15),
      lineheight = 1.1
    ),
    plot.caption = element_markdown(
      size = 9,
      color = text_col,
      margin = margin(t = 15),
      lineheight = 1.1
    ),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
    axis.title.x = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(
      color = "gray90",
      linewidth = 0.15
    ),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

```

## 4 Save plot

```{r}
#| label: Save plot
#| eval: false

ggsave(
  filename = "likely.png", 
  plot = p,
  width = 7, 
  height = 8,
  dpi = 300, 
  bg = "white"
)
```



